\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Assessing economic inequality with tax data - Switzerland from 1945 to 2010}

### 2014-06-09 paper/subdocuments/data_and_methods.Rnw
citations and spell check

Studies on inequality have to adress several thorny challenges. It starts with answering three crucial questions: First of all one has to define, which concepts should to be looked at. This refers to answering the question about inequality of what. Secondly one has to be clear about the unit of analysis. This refers to answering the question about inequality among whom. Thirdly, one has to choose an appropriate measure of inequality. All this questions are ideally answered considering theory and a given research question. Often it has to be answered in context of a given dataset. Therefore we start this section with a description of the FTA-Tax Data. Based on a review on the literature about the measurement concepts in an ideal world, we discuss  the advantages and shortcomings of tax data compared to other data sources - namely survey data. Along with the despriction we explain the methods and thechnics we use to construct time-series of inequality-measure for income for Switzerland, which we will present in the result sections. By doing this, we demonstrate the suitability of the FTA Tax Data to assess the development of inequality in Switzerland

\subsection{Tax statistics in Swizterland}

@@

Most studies on inequality focus on income inequality solely. However, recent activities emphasize the need of a broader conceptualization. A recent publication from the oecd (2013) condense these ideas into the ICW framework (income, consumption and wealth), which is ment to be an internationally agreed framework on micro-level statistics.\footnote{Harmonisation with other international standards was an important objective that guided the work of the Expert Group in developing the ICW Framewrok presented in this publication. Considered main stadards were the System of National Accounts (SNA 2008), the canberra Group Handbook on Household Income Statistics (United Nations 2011), the final report of the 17th International Conference of Labour Statisticians (ILO 2004) and the UNECE/CES recommendations for the 2010 Censuses of Population and Housing (UNECE & Eurostat 2006)} According to the framework it is best to look at income, consumption and wealth as three separate but interrelated dimensions of people's economic well-being. To gain policy relevant insight, it is recomended to look at the distribution of all three distributions simultaneously. Some households with low income, for example, may report adequate levels of consumption expenditure or welth holdings, or vice-versa. But it is also stated (OED 2013:18):"[...] integrated analysis at the household level has significant data requirements that go beyond the measurement efforts currently undertaken in most countries."\footnote{The Luxembourg Wealth Study Database is currently facing this shortcomings by colecting and provding a database following ths broader concept of economic well-being. http://www.lisdatacenter.org/our-data/lws-database/} \\

This last statment holds for Swizterland too, although the HBS study is strongly influced by the recommendations of the Canberra group handbook (United Nations, 2011), which concepts are part of the ICW framework. Albeit the awarness of an assessment of income, consumption and wealth simuntanously is rising, we  focus our analysis on income, which is undoubtly a crucial indicator of economic wellbeeing. However, the Federal Tax Office publishes statistics on income and wealth . But it is not possible to analyse the jointly distribution on the individual or household level. Also measures of consumption are largely missing in tax data, albeit deductions can be understood as mandatory consumptions somehow.

@@

The assessment of income inequality is influenced by the definition of the income itself. Marketincome or disposable income for example differ by substantial meaning and by the expected degree of inequality. Therefore the awarness of the analyzed concept is crucial. Terminology can sligthly differ, while common concepts can be identified (for detailed discussion see: OECD (2013:44), Canberra group (2011:24)). Figure xy shows a stylized framework, which includes a distinction of common income sources \footnote{Income from production of household services for own consumption is excluded because this income is hard to measure and not covered in the FTA tax data} and shows the central steps of redistribution. Within this framework common income definitions are situated. These definitions are contrasted with income concepts, which can be derived out of the FTA Tax data.

\includegraphics[width=\maxwidth]{figure/income_definition}

@@

The central income according to tax data is the taxable income. It includes all reported incomes (income from employment, income from property and received transfers\footnote{Mean-tested benefits are not taxed and therefore not included in Tax data. Income for low income groups are therefore underestimated.}) minus several deductions. It is therefore neither a pre-transfer income  nor a post-transfer income measure. It's rather something inbetween. As the FTA tax statistitcs include some but not all deductions\footnote{The difference between the real Total income and the taxable income are deductions. This includes: professional expenses, travel expenses, interest on debt, Alimonies, Training costs, two-earner deduction, Party Contributions, Payment into pillar 3a, purchases in the pension fund and sideline deductions} it is possible to calculate a sort of "total income". As some deducations can be interpreted as compulsary expenses similar to taxes the step towards Total income is a step away from the income, which can be used for consumption. Similiar when calculating the disposable income out of the taxable income through accouting the reported federal taxes, this is a step towards the income, which is left in the basket for consumption (disposable income). Again it is not a "pure" disposable income, because cantonal, municipal taxes and taxes from churches, which represent the bulk ot taxes in Switzerland, are missing.   \\

@@

The agreed standard on the statistical unit, which should be the base of inequality analysis, are households not individuals (OECD 2013:60). Indeed it are the individuals, who receive income, own wealth and experience economic wellbeing, but their possibility to do so, is strongly tied to the concept of household. This comprises all persons under the same housing arrangement. The basic underlying assumption for colection data on households level insted of individual level is, that people in the same household share ressources and therefore pool their incomes (when two or more earners live together) and/or use the household income to provide the essentials of living for every householdmember (also non-earing members, like children). Additionaly, there are economies of scale when people share living space and commodities and they therefore benefit from the sharing. The compare the individual economic well-beeing among individuals living in different households usualy equivalence scales are used (see OECD 2013,173, Buhmann et al., 1998). \\

@@

In tax data, however, the units are represented according to administrative rules. Tax units therefore neither represent individuals in every case nor true households. Tax units rather represent individuals and couples, but couples, who are married or officially registrered. This doesn't imply, that those couples live together, as it is needed to satisfy the definition of a household. On the other hand, is it quit likely that more than one tax units live in the same household (unmarried/unregistered couples, see M?ller and Schoch (2014:99)). It is therefore not directly possible to elicit households and household inomce out of tax data. This might influence the assessment inequality, taking into account the change from traditional household and family structures over the last century. This is an inequality related issue, where relevant studies are missing.

@@

Often inequality is assessed on national level, which implies, that studies try to cover the whole population of the country of interest. This is a special thorny task for surveys working with samples, because nonresponse is a major source of bias (Bethlehem et al., 2011).  Korinek, Mistiaen and Ravallion (2006) show, that the position in the income distribution influence the probability to participate in a survey. Low income and high income households are more likely to refuse survey response, which leads to an overrepresentation of middle income households. This process can be referred to as the "middleclass bias" Diekmann (2009). Missing data in household surveys is therefore not missing at random, which has an impact on the measures of inequality. The magnitude of this bias in Switzerland, however, is unknown. Strategies to handle this kind of bias are discussed in the literature (S?rndal et al. 1992), but require a register for every unit, that is proportional to income. Currently there exists no such register for Switzerland (ecoplan 2014:43). Currently used microdata sets, which are used for official publications concerning inequality in Switerzland (SILC and HABE) are furthermore confronted with a constructed coverage problems, because this surveys rely on phoneregister, which excludes the households not having a registered connection. \\
%(Problems of collective households, ecoplan 2014:45). Erw?hnen?

The issue of incomplete coverage is far less dramatical with tax data. Essentially every permanent resident in Switzerland, who has completed the age of 18 years (respectivley 20 years prior to 1996) is taxed on a yearly base (or every two years before the change of the tax system). In principle this leads to a full representation of the adult population of Switzerland and a complete coverage of the income distrubtion. This includes a separation of normal cases, which embrace the majority of taxpayers, and the special cases, which cover several unusual situations. Most important high net wealth individuals taxed according to their expenditures.However, special attention has to be paid to tax units with none or very low incomes. Even though they have to hand in a tax return, their income do not show up in the statistics if their income after deductions falls below 15'000 CHF and they are therefore not taxed with direct federal taxes. This is possible for normal and special cases alike. From 1995/1996 until 2010 the number of non-taxed units are reported, but not for the years before.

@@

\emph{Measuring inequality or concentration} \\
To be able to make qualifying statements about a distribution or to compare different distributions, the concept of inequality turned out to be the most appropriate and thus the most commonly used dimension. Vgl. Allison (1978), Engelhardt (2000), Cowell (2000) oder Hao/Naiman (2010)
%[Zusammenfassung bestehender Überlegung und Hinleitung zu den Masszahlen, die wir verwenden werden (vermutlich Gini und relative distribution methods) \\

neuere Indikatoren; Polaritätsindex: relative distribution methods in the Social Sciences (Handcock/Morris 1999) \\

Gini is mainly used for international comparison. However, other measures are often reported along (gini is more sensitive to changes in the middle of the distribution than to changes in the tails (OECD, 2008:37). Newer branches of inequality studies emphasize the need for broader measures of inequality, which allow better analyses about the change of inequality and namely statements about the area of change (downgrading/upgrading) (Alderson and Doran 2013). \\

Alvaredo (2010) shows formally, how top incomes shares are related to the gini-coefficient. Following his argument, it is crucial to consider top incomes. Leigh (2007) also analyzed the relationship between the Gini (and further inequality measures) and top income shares in a panel of 13 countries (Switzerland is part of the studie). 
%[He also discusses tax data studies in general-> have a closer look] \\
%[We definitely have to argue, why it isn’t sufficient to report top income shares, because if it is, there is no need to have another study about Switzerland (national level>Martinez, cantonal level> gorgas and Schaltenegger)]

@@

\textbf{Summary of tax data drawbacks}
\begin{itemize}
\item Misreporting of incomes (high earners have an interest in getting tax beneficial ``income'' see above)
\item with aggregated data the adjustment for household members is not possible (no equivalization possible). This is a point, which is hardly mentioned in the ``top-income'' literature, albeit it is a crucial issue in ``theorized'' concepts of income. To generalize one can say, that the concept of tax units (individuals and couples + no direct information about household members at least not in the aggregated tax tables) is not congruent with the concept of household. This might influence the overall inequality, taking into account the change from traditional household and family structures over the last century. This is an inequality related issue, where relevant studies are missing. What are the assumptions?
\item the tax data income definition summarize labor income, capital income and taxable transfer payments (no means-tested benefits as welfare aid). Distinction between the different income sources is not possible, albeit the mechanisms of changes in the specific distribution can be very different.
\item Some tax units are not included in the tax tables (see Foellmi and Martinez 2013:11f).(a) only taxed cases are included (partly a problem, for certain periodes we know how many cases are excluded because of low or none income)	- (b) cases tax at source (Quellenbesteuerung). Foreign nationals living in Switzerland but with a yearly or any other temporary resident permit only. In some border-cantons the share of this group is very high (20\%) 
%Can we ignore this)
(c) staff of international organizations based in Switzerland	- (d) non-fillers show up in the tables as long as they are registred. This persons get an imputed income (older tax tax return and information given by employers. Not registered non-	fillers are not in the records.
- (e) tax evasion. Feld and Frey (2007) report about tax evasion in Switzerland, it should be somewhat above 20 percent on average. With the (strong) assumption that the pattern of tax evasion over time is stable, this is a minor problem for inequaly measures over time.
\end{itemize}

@@

Atkinson et al. (2009) estimate that CPS survey data fail to capture about half of the overall increase in inequality measured by the Gini coefficient, a result confirmed by Alvaredo (2010). \\

See for other countries: Siminski et al 2003 (Australia), Brewer et al 2008, UK, Burkhauser et al. 2009, US)
%	•	What about the special cases? Foelmi and Martinez argue, that it is crucial to include the special cases because they include the high net wealth individuals taxed according to their expenditures. Exclusion leads to an underestimation of inequality
%	•	Capital gains? Should not be included 
The discussion about problems with reporting income is fairly exhaustive. What about wealth?

It is well recognized that the traditional sources of wealth distribution data are unlikely to provide an accurate picture of wealth ownership in the top-tail of the distribution. Credit Suisse (2011) makes use of the information in the ``Rich Lists'' published by Forbes Magazine to adjust the wealth distribution pattern in the highest wealth ranges. \\

Additonaly the Credit Suisse Reports states (2011:8) that these data my be less subject to response bias, but my be more prone to valuation problems, especially in connection with pension assets and debts.

@@

\textbf{Incomplete coverage of the population (left censored data.)} What can be done about the not-taxed? Del et al. (2007) impute for non-fillers the 20 percentage of the annual average income. This flattens the distribution on the left side, which is not a problem if you are interested in the top income shares, but it would surly affect overall measures of inequality. Furthermore Del et al. calculate the proportion of non-fillers by estimating the total of tax units out of the population records. \\

\textbf{changes in taxation system  (switch from annual to biannual taxation)} In the mid-1990s a fundamental change in the Swiss tax system took place by switching form the two-years based praenumerando taxation to the one-year based postnumerando taxation. This change was enacted with a transitional period of several years, during which each canton could choose when to adopt the new system.  This is why during the transitional period from 1995 to 2003 there is no uniform tax data published on the Swiss level but only data on the cantonal level  (Foellmi and Martinez:8f). \\
%es wird erwartet, dass der Wechsel Ungleichheitsmasse beeinflussen. Yearly fluctuations are dampened, when income is measured on a two-yearly basis.

@@

\textbf{deductions} Del et al. (2007:477):" we can check with statistics for 1971-72 (as well as later years) presented both by size of income before deductions and income after deductions that adding back deductions does not introduce any significant error in our estimates."
Gorgas and Schaltenegger (2011:5): ``..., information on [...] deductions is provided in the tax statistics, thus, we could add the personal deductions to the income data to obtain a consistent series over time''. Können wir das auch? Zumindest für gewisse Zeiträume? Das wäre noch gut. \\

Studies on income try to focus on the disposable income, which subtracts certain expenditures from the primary income. Deductions reflect somehow compulsory expenditures and thus taxable income can be seen as a sort of pseudo disposable income. On the other hand deductions can affect the distribution. There are recent studies about the correlation of progressivity and deductions in Switzerland, which examines if deductions have a ``perverse redistribution'' effect by redistributing income from the lower middle class to the upper middle class (vgl. Peters 2011 and Interpellation Barbara Gysel (2009).

@@

\begin{enumerate}
\item There is a small overall upward trend which we assume to be the Federal Administrations inflation adjustments to the tax threshold.
\item Geneva and Tessin show wild changes but those might be explained by the tax gap ("Bemessungslücke") that people exploited when the cantons changed the tax system. It remains unclear however why we can't see similar patter within other cantons.
\item There is some variance and we see different patterns over time and cantons. When estimating gini coefficients or the like we must therefore assume that ignoring ``the zeros'' leads to a bias that is not stable over time.
\end{enumerate}

@@

\begin{enumerate}
\item Add the zeros as a seperate group
\item Fit a model to predict the inequality measure (e.g. the gini coefficient) using the share of zeros as a predictor
\end{enumerate}

By adding the zeros as a seperate group we face several problems. With the exception of Geneva 1995/96 Pareto interpolation of percentiles p20 and above seems viable. Low percentiles however would need to be extrapolated so we might impose unrealistic assuptioms. Even when estimating ``safe'' measures above p20 (better p50) or a gini coefficient we need to make an assumption about the income structure of that group (a distribution, a mean income or zero-income). To be consistent with the measure of ``taxable income'' we could assume an income of zero for that group, calculate gini coefficients and compare them with the original (uncorrected/plain) version. To get a rough number we could (cantonwise) check the squared correlation between plain and corrected gini.

@@

In our case we can clearly reject the hypothesis that all interactions are zero ($p=\Sexpr{round(ftest$p,3)}$). This leads to the conclusion  that gini coefficients are biased by the variation of the zero-share which is kind of obvious but at the same time we can use the model to report adjusted gini coefficients. For example one might be interested in how inequality would had developed if the zero-share would have been constant over time. (Note RF: predict all data point using canton, time and the initial OR final zero-share to homogenize the time series)
Furthermore we can quantify how large the bias is and we can do this seperately for tax periods or seperately for cantons.

@@

We can see the the model fit reduces to explaining 61.5\% of the gini variation versus 92.6\% when the information from the zero-shares was used. Although this so some extent attributable to the additional 26 paramters: this is huge.

@@

A positive coefficient (e.g. Schwyz) can be read as follows: In periods with many zeros we measure higher gini coefficients. We can derive, that the distribution of income is more scewed for high incomes than for low incomes. Simply speaking, the contrast between low and middle class is less pronounced than the contrast between middle und high class. One possible explanation would be that incomes stem from two different populations: 1) the people of Schwyz who possibly follow a log-normal or gamma distribution and 2) particularly rich people who moved to Schwyz to avoid taxes.

A negative coefficient (e.g. Geneva) means the more zeros there are the smaller the gini measure was compared to other tax periods within that canton (remember this is a fixed-effects model). This is the case we would usually expect: more zeros mask inequality that arises from the bottom.

What can we conclude from that analysis? First one must notice that aggregate measure like the gini (or others) do not always react in the same way when we cut off one part of the distribution, therefore the measures calculated from tax data is biased. On the other hand, the model coefficient of Switzerland as a whole is not significant suggesting that the cantonal biases cancel out each other. This seems plausible. Most of the ``tax optimization'' happens within Switzerland so the rich people who moved to Schwyz are now missing at another canton.

>>>>

Studies on inequality have to address several thorny challenges. It starts with answering three crucial questions: First of all one has to define, which concepts should to be looked at. This refers to answering the question about inequality of what. Secondly one has to be clear about the unit of analysis. This refers to answering the question about inequality among whom. Thirdly, one has to choose an appropriate measure of inequality. All this questions are ideally answered considering theory and a given research question. Often it has to be answered in context of a given dataset. Therefore we start this section with a description of the FTA-Tax Data. Based on a review on the literature about the measurement concepts in an ideal world, we discuss  the advantages and shortcomings of tax data compared to other data sources - namely survey data. Along with the description we explain the methods and techniques we use to construct time-series of inequality-measure for income for Switzerland, which we will present in the result sections. By doing this, we demonstrate the suitability of the FTA Tax Data to assess the development of inequality in Switzerland

\subsection{Tax statistics in Switzerland}

@@

Most studies on inequality focus on income inequality solely. However, recent activities emphasize the need of a broader conceptualization. A recent publication from the \citet{oecd_oecd_2013 UNSICHER} condense these ideas into the ICW framework (income, consumption and wealth), which is ment to be an internationally agreed framework on micro-level statistics.\footnote{Harmonization with other international standards was an important objective that guided the work of the Expert Group in developing the ICW Framework presented in this publication. Considered main standards were the System of National Accounts \citep{sna_system_2008}, the Canberra Group Handbook on Household Income Statistics \citep{united_nations_canberra_2011}, the final report of the 17th International Conference of Labour Statisticians \citep{international_labour_organisation_ilo_final_???? UNSICHER}} and the UNECE/CES recommendations for the 2010 Censuses of Population and Housing \citep{unece_conference_2006}} According to the framework it is best to look at income, consumption and wealth as three separate but interrelated dimensions of people's economic well-being. To gain policy relevant insight, it is recommended to look at the distribution of all three distributions simultaneously. Some households with low income, for example, may report adequate levels of consumption expenditure or wealth holdings, or vice-versa. But it is also stated \citep[18]{oecd_oecd_2013 UNSICHER}:"[...] integrated analysis at the household level has significant data requirements that go beyond the measurement efforts currently undertaken in most countries."\footnote{The Luxembourg Wealth Study Database is currently facing this shortcomings by collecting and providing a database following this broader concept of economic well-being. http://www.lisdatacenter.org/our-data/lws-database/} \\

This last statement holds for Switzerland too, although the HBS study is strongly influenced by the recommendations of the Canberra group handbook (United Nations, 2011), which concepts are part of the ICW framework. Albeit the awareness of an assessment of income, consumption and wealth simultaneously is rising, we  focus our analysis on income, which is undoubtedly a crucial indicator of economic wellbeing. However, the Federal Tax Office publishes statistics on income and wealth . But it is not possible to analyze the jointly distribution on the individual or household level. Also measures of consumption are largely missing in tax data, albeit deductions can be understood as mandatory consumptions somehow.

@@

The assessment of income inequality is influenced by the definition of the income itself. Market income or disposable income for example differ by substantial meaning and by the expected degree of inequality. Therefore the awareness of the analyzed concept is crucial. Terminology can slightly differ, while common concepts can be identified (for detailed discussion see: \citet[44]{oecd_oecd_2013 UNSICHER}, \citet[24]{united_nations_canberra_2011}). Figure~\ref{fig:incdef} shows a stylized framework, which includes a distinction of common income sources \footnote{Income from production of household services for own consumption is excluded because this income is hard to measure and not covered in the FTA tax data} and shows the central steps of redistribution. Within this framework common income definitions are situated. These definitions are contrasted with income concepts, which can be derived out of the FTA Tax data.

\begin{figure}
\centering
\includegraphics[width=\maxwidth]{figure/income_definition}
\caption{Income definitions}
\label{fig:incdef}
\end{figure}

@@

The central income according to tax data is the taxable income. It includes all reported incomes (income from employment, income from property and received transfers\footnote{Mean-tested benefits are not taxed and therefore not included in Tax data. Income for low income groups are therefore underestimated.}) minus several deductions. It is therefore neither a pre-transfer income  nor a post-transfer income measure. It's rather something in between. As the FTA tax statistics include some but not all deductions\footnote{The difference between the real Total income and the taxable income are deductions. This includes: professional expenses, travel expenses, interest on debt, Alimonies, Training costs, two-earner deduction, Party Contributions, Payment into pillar 3a, purchases in the pension fund and sideline deductions} it is possible to calculate a sort of "total income". As some deductions can be interpreted as compulsory expenses similar to taxes the step towards Total income is a step away from the income, which can be used for consumption. Similar when calculating the disposable income out of the taxable income through accounting the reported federal taxes, this is a step towards the income, which is left in the basket for consumption (disposable income). Again it is not a "pure" disposable income, because cantonal, municipal taxes and taxes from churches, which represent the bulk of taxes in Switzerland, are missing.   \\

@@

The agreed standard on the statistical unit, which should be the base of inequality analysis, are households not individuals \citep[60]{oecd_oecd_2013 UNSICHER}. Indeed it are the individuals, who receive income, own wealth and experience economic wellbeing, but their possibility to do so, is strongly tied to the concept of household. This comprises all persons under the same housing arrangement. The basic underlying assumption for collection data on households level instead of individual level is, that people in the same household share resources and therefore pool their incomes (when two or more earners live together) and/or use the household income to provide the essentials of living for every household member (also non-earing members, like children). Additionally, there are economies of scale when people share living space and commodities and they therefore benefit from the sharing. The compare the individual economic well-being among individuals living in different households usually equivalence scales are used (see \citealt[173]{oecd_oecd_2013 UNSICHER}, \citealt{Buhmann et al., 1998 FEHLT}). \\

@@

In tax data, however, the units are represented according to administrative rules. Tax units therefore neither represent individuals in every case nor true households. Tax units rather represent individuals and couples, but couples, who are married or officially registered. This doesn't imply, that those couples live together, as it is needed to satisfy the definition of a household. On the other hand, is it quit likely that more than one tax units live in the same household (unmarried/unregistered couples, see \citet[99]{muller_vermogenslage_2014}). It is therefore not directly possible to elicit households and household income out of tax data. This might influence the assessment inequality, taking into account the change from traditional household and family structures over the last century. This is an inequality related issue, where relevant studies are missing.

@@

Often inequality is assessed on national level, which implies, that studies try to cover the whole population of the country of interest. This is a special thorny task for surveys working with samples, because nonresponse is a major source of bias \citep{bethlehem_handbook_2011}.  \citep{korinek_survey_2006} show, that the position in the income distribution influence the probability to participate in a survey. Low income and high income households are more likely to refuse survey response, which leads to an overrepresentation of middle income households. This process can be referred to as the "middleclass bias" \citep{diekmann_empirische_2009}. Missing data in household surveys is therefore not missing at random, which has an impact on the measures of inequality. The magnitude of this bias in Switzerland, however, is unknown. Strategies to handle this kind of bias are discussed in the literature \citep{S?rndal et al. 1992 FEHLT}, but require a register for every unit, that is proportional to income. Currently there exists no such register for Switzerland \citep{ecoplan 2014:43 FEHLT}. Currently used micro datasets, which are used for official publications concerning inequality in Switzerland (SILC and HABE) are furthermore confronted with a constructed coverage problem, because these surveys rely on the phone register, which excludes the households not having a registered connection. \\
%(Problems of collective households, ecoplan 2014:45). Erw?hnen?

The issue of incomplete coverage is far less dramatical with tax data. Essentially every permanent resident in Switzerland, who has completed the age of 18 years (respectively 20 years prior to 1996) is taxed on a yearly base (or every two years before the change of the tax system). In principle this leads to a full representation of the adult population of Switzerland and a complete coverage of the income distribution. This includes a separation of normal cases, which embrace the majority of taxpayers, and the special cases, which cover several unusual situations. Most important high net wealth individuals taxed according to their expenditures.However, special attention has to be paid to tax units with none or very low incomes. Even though they have to hand in a tax return, their income do not show up in the statistics if their income after deductions falls below 15'000 CHF and they are therefore not taxed with direct federal taxes. This is possible for normal and special cases alike. From 1995/1996 until 2010 the number of non-taxed units are reported, but not for the years before.

@@

\emph{Measuring inequality or concentration} \\
To be able to make qualifying statements about a distribution or to compare different distributions, the concept of inequality turned out to be the most appropriate and thus the most commonly used dimension. See \citet{Allison 1978 FEHLT}, \citet{engelhardt_modelle_2000}, \citet{cowell_chapter_2000} and \citet{hao_assessing_2010}.
%[Zusammenfassung bestehender Überlegung und Hinleitung zu den Masszahlen, die wir verwenden werden (vermutlich Gini und relative distribution methods) \\

neuere Indikatoren; Polaritätsindex: relative distribution methods in the Social Sciences \citep{handcock_relative_1999} \\

The Gini coefficient is mainly used for international comparison. However, other measures are often reported along because the Gini coefficient is more sensitive to changes in the middle of the distribution than to changes in the tails \citep[37]{OECD, 2008 FEHLT}. Newer branches of inequality studies emphasize the need for broader measures of inequality, which allow better analyses about the change of inequality and namely statements about the area of change (downgrading/upgrading) \citep{Alderson and Doran 2013 FEHLT}. \\

\citet{Alvaredo (2010) FEHLT} shows formally, how top incomes shares are related to the Gini-coefficient. Following his argument, it is crucial to consider top incomes. \citet{leigh_how_2007} also analyzed the relationship between the Gini (and further inequality measures) and top income shares in a panel of 13 countries (Switzerland is part of the study). 
%[He also discusses tax data studies in general-> have a closer look] \\
%[We definitely have to argue, why it isn’t sufficient to report top income shares, because if it is, there is no need to have another study about Switzerland (national level>Martinez, cantonal level> gorgas and Schaltenegger)]

@@

\textbf{Summary of tax data drawbacks}
\begin{itemize}
\item Misreporting of incomes (high earners have an interest in getting tax beneficial ``income'' see above)
\item with aggregated data the adjustment for household members is not possible (no equivalization possible). This is a point, which is hardly mentioned in the ``top-income'' literature, albeit it is a crucial issue in ``theorized'' concepts of income. To generalize one can say, that the concept of tax units (individuals and couples + no direct information about household members at least not in the aggregated tax tables) is not congruent with the concept of household. This might influence the overall inequality, taking into account the change from traditional household and family structures over the last century. This is an inequality related issue, where relevant studies are missing. What are the assumptions?
\item the tax data income definition summarize labor income, capital income and taxable transfer payments (no means-tested benefits as welfare aid). Distinction between the different income sources is not possible, albeit the mechanisms of changes in the specific distribution can be very different.
\item Some tax units are not included in the tax tables (see \citealt[11f]{foellmi_volatile_2013}).(a) only taxed cases are included (partly a problem, for certain periods we know how many cases are excluded because of low or none income)	- (b) cases tax at source (Quellenbesteuerung). Foreign nationals living in Switzerland but with a yearly or any other temporary resident permit only. In some border-cantons the share of this group is very high (20\%) 
%Can we ignore this)
(c) staff of international organizations based in Switzerland	- (d) non-fillers show up in the tables as long as they are registered. This persons get an imputed income (older tax tax return and information given by employers. Not registered non-	fillers are not in the records.
- (e) tax evasion. \citet{feld_tax_2006 UNSICHER weil vorher 2007 stand} report about tax evasion in Switzerland, it should be somewhat above 20 percent on average. With the (strong) assumption that the pattern of tax evasion over time is stable, this is a minor problem for inequality measures over time.
\end{itemize}

@@

\citet{atkinson_top_2009} estimate that CPS survey data fail to capture about half of the overall increase in inequality measured by the Gini coefficient, a result confirmed by \citet{Alvaredo (2010) FEHLT}. \\

See for other countries: \citealt{Siminski et al 2003 FEHLT} (Australia), \citealt{Brewer et al 2008 FEHLT}, UK, \citealt{Burkhauser et al. 2009 FEHLT}, US)
%	•	What about the special cases? Foelmi and Martinez argue, that it is crucial to include the special cases because they include the high net wealth individuals taxed according to their expenditures. Exclusion leads to an underestimation of inequality
%	•	Capital gains? Should not be included 
The discussion about problems with reporting income is fairly exhaustive. What about wealth?

It is well recognized that the traditional sources of wealth distribution data are unlikely to provide an accurate picture of wealth ownership in the top-tail of the distribution. \citet{credit_suisse_global_2011} makes use of the information in the ``Rich Lists'' published by Forbes Magazine to adjust the wealth distribution pattern in the highest wealth ranges. \\

Additionally the Credit Suisse Reports states that these data my be less subject to response bias, but my be more prone to valuation problems, especially in connection with pension assets and debts \citep[8]{credit_suisse_global_2011}.

@@

\textbf{Incomplete coverage of the population (left censored data.)} What can be done about the not-taxed? \citet{dell_income_2007} impute for non-fillers the 20 percentage of the annual average income. This flattens the distribution on the left side, which is not a problem if you are interested in the top income shares, but it would surly affect overall measures of inequality. Furthermore the authors calculate the proportion of non-fillers by estimating the total of tax units out of the population records. \\

\textbf{changes in taxation system  (switch from annual to biannual taxation)} In the mid-1990s a fundamental change in the Swiss tax system took place by switching form the two-years based praenumerando taxation to the one-year based postnumerando taxation. This change was enacted with a transitional period of several years, during which each canton could choose when to adopt the new system.  This is why during the transitional period from 1995 to 2003 there is no uniform tax data published on the Swiss level but only data on the cantonal level \citep[8f]{foellmi_volatile_2013}. \\
%es wird erwartet, dass der Wechsel Ungleichheitsmasse beeinflussen. Yearly fluctuations are dampened, when income is measured on a two-yearly basis.

@@

\textbf{deductions} \citet[477]{dell_income_2007}:" we can check with statistics for 1971-72 (as well as later years) presented both by size of income before deductions and income after deductions that adding back deductions does not introduce any significant error in our estimates."
\citet[5]{schaltegger_evolution_2011 UNSICHER}: ``..., information on [...] deductions is provided in the tax statistics, thus, we could add the personal deductions to the income data to obtain a consistent series over time''. Können wir das auch? Zumindest für gewisse Zeiträume? Das wäre noch gut. \\

Studies on income try to focus on the disposable income, which subtracts certain expenditures from the primary income. Deductions reflect somehow compulsory expenditures and thus taxable income can be seen as a sort of pseudo disposable income. On the other hand deductions can affect the distribution. There are recent studies about the correlation of progressivity and deductions in Switzerland, which examines if deductions have a ``perverse redistribution'' effect by redistributing income from the lower middle class to the upper middle class (see \citealt{peters_steuerabzuge:_2011} and \citet{Interpellation Barbara Gysel (2009) FEHLT. das interpellation barbara sieht falsch aus}.

@@

\begin{enumerate}
\item There is a small overall upward trend which we assume to be the Federal Administrations inflation adjustments to the tax threshold.
\item Geneva and Tessin show wild changes but those might be explained by the tax gap ("Bemessungslücke") that people exploited when the cantons changed the tax system. It remains unclear however why we can't see similar patter within other cantons.
\item There is some variance and we see different patterns over time and cantons. When estimating Gini coefficients or the like we must therefore assume that ignoring ``the zeros'' leads to a bias that is not stable over time.
\end{enumerate}

@@

\begin{enumerate}
\item Add the zeros as a separate group
\item Fit a model to predict the inequality measure (e.g. the Gini coefficient) using the share of zeros as a predictor
\end{enumerate}

By adding the zeros as a separate group we face several problems. With the exception of Geneva 1995/96 Pareto interpolation of percentiles p20 and above seems viable. Low percentiles however would need to be extrapolated so we might impose unrealistic assumptions. Even when estimating ``safe'' measures above p20 (better p50) or a Gini coefficient we need to make an assumption about the income structure of that group (a distribution, a mean income or zero-income). To be consistent with the measure of ``taxable income'' we could assume an income of zero for that group, calculate Gini coefficients and compare them with the original (uncorrected/plain) version. To get a rough number we could (by cantons) check the squared correlation between plain and corrected Gini.

@@

In our case we can clearly reject the hypothesis that all interactions are zero ($p=\Sexpr{round(ftest$p,3)}$). This leads to the conclusion  that Gini coefficients are biased by the variation of the zero-share which is kind of obvious but at the same time we can use the model to report adjusted Gini coefficients. For example one might be interested in how inequality would had developed if the zero-share would have been constant over time. (Note RF: predict all data point using canton, time and the initial OR final zero-share to homogenize the time series)
Furthermore we can quantify how large the bias is and we can do this separately for tax periods or separately for cantons.

@@

We can see the the model fit reduces to explaining 61.5\% of the Gini variation versus 92.6\% when the information from the zero-shares was used. Although this so some extent attributable to the additional 26 parameters: this is huge.

@@

A positive coefficient (e.g. Schwyz) can be read as follows: In periods with many zeros we measure higher Gini coefficients. We can derive, that the distribution of income is more skewed for high incomes than for low incomes. Simply speaking, the contrast between low and middle class is less pronounced than the contrast between middle and top class. One possible explanation would be that incomes stem from two different populations: 1) the people of Schwyz who possibly follow a log-normal or gamma distribution and 2) particularly rich people who moved to Schwyz to avoid taxes.

A negative coefficient (e.g. Geneva) means the more zeros there are the smaller the Gini measure was compared to other tax periods within that canton (remember this is a fixed-effects model). This is the case we would usually expect: more zeros mask inequality that arises from the bottom.

What can we conclude from that analysis? First one must notice that aggregate measure like the Gini (or others) do not always react in the same way when we cut off one part of the distribution, therefore the measures calculated from tax data is biased. On the other hand, the model coefficient of Switzerland as a whole is not significant suggesting that the cantonal biases cancel out each other. This seems plausible. Most of the ``tax optimization'' happens within Switzerland so the rich people who moved to Schwyz are now missing at another canton.
